# Introduction {#intro}

Welcome to our open-source and cross-listed course on Data Science for kids who can't read data but want to learn to read data and other things too. in R.


We hope your knowledge of R increases, just like the graph below!
  * (This is just making up reasons for the sample plot)

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].


Each `.Rmd` creates a unique **CHAPTER**. 
  * Organizing by content probably makes the most sense
  * Gives more flexibility to adapt 2/3 class days.
  * *other thoughts??*
  
