

# Lesson 3: Comparing Data


**NEEDS A LOT OF WORK**


## Data Distributions

### Normal Distributions

First we'll generate a normal distribution with the `rnorm()` function. This takes 3 arguments: `n, mean, sd`, which you can see filled in below. While we could print out a list of all these values, it's not easy to *understand* a list of numbers
```{r}

normal_dist <- rnorm(n = 100, ## 100 samples
                     mean = 10, ## with a mean of 10
                     sd = 1 ## and a standard deviation of 1
                     )


normal_dist
```


Another better way to look at data would be to **visualize** or **plot** it. One way to to that is with a **histogram**, which groups **continuous values** into **bins**, then plots the **frequency** for each bin. 

In R, we use the `hist()` function to plot a histogram of data. We can (try to) control the number of bins with the `breaks` argument, but note that it doesn't always match up. The `hist()` function will adjust based on the distribution of the data.

```{r norm_dist_plot}
hist(normal_dist,breaks = 5)
```

Another way to visualize this would be with a d

### What *is* normal?
#### Quantitative summaries

5num summary
* Min, 25th percentile, median, 75th percentile, Max

```{r}
tab_normal_dist <- summary(normal_dist)

```


We can print the table in R by calling its name.

```{r}
tab_normal_dist
```

Mean, standard deviation

#### Meaningful Comparisons

How to compare apples to oranges? Standardize the units / standardize the data


```{r}
data1 <- rnorm(n=1000, 
              mean = 100,
              sd = 10)

data2 <- rnorm(n=1000,
               mean = 60, 
               sd = 25)

```
Are these the same distribution?

Any issues??

```{r, rnorm_hist_comp}
layout(matrix(1:2, ncol = 2))
hist(data1)
hist(data2)
```

```{r}
total_range <- range(data1, data2)
```


Are they the same?

```{r, rnorm_hist_comp_true}
layout(matrix(1:2, ncol = 2))
hist(data1, xlim = total_range)
hist(data2, xlim = total_range)
```

Numerically / tabularly

Often times its important to tables of **summary statistics**

```{r}

norm_comp_tab <- rbind(summary(data1),
                       summary(data2))

norm_comp_tab

```


Making the table a little nicer. Also an example of **conditional programming**.


```{r}


rownames(norm_comp_tab) ## they're null


if(is.null(rownames(norm_comp_tab))){
  rownames(norm_comp_tab) <- c("data1", "data2")
}

```
When working with **Rmarkdown** we can take advantage of `knitr` and `pandoc` to nice looking tables even easier.


```{r norm_summary_tab}

knitr::kable(norm_comp_tab)
```
**How** transform the data

Simple transformation (multiply all values by 100)
  * to convert units
  * other examples?

Complex transformations
  * log-transformation (*DEE: not a fan*)
  * z-scores (*DEE: a better option*)

**Why** transform the data?
  * Real world applications?
  * Is it always appropriate to transform data?

  
### Skews

What to do if the data are **not** normal?

##  Statisitcal testing of simple data sets

### t-tests, ANOVA, chi2

## Relationships between variables in simple data sets
### Correlation, Linear Regression
#### Simple LM
#### Complex LM

### Genearlized Linear Model


For now, I have 3 main chapters for each of the main sections: 
  * Basics of data science / R \@ref(unit1) 
  * Applications/critiques using IPUMS data \@ref(unit2)
  * Student-driven projects \@ref(unit3)

Each of these **Chapters** contains multiple sections. We'll likely want to break these sections out into their own `.Rmd` files as they get fleshed out. For now, I'll try to keep the abundance of files limited.

**NOTE:** As these actually get filled out, we will probably want to insert different `part`s to the book (EG, the content of Unit 1 is covered in `Part I`).
  * Declare parts with `# (PART) Part I {-}` immediately before the first chapter `#` it contains.
  
**Topics to include:**
  * What is data?
  * Everything can be data
  * How do we interpret data
  * Tables
  * Plots
  * Univariate distributions
  * What can they tell us
  * Multi-modality in distributions
  * Categorical vs continuous data
  * Don’t need to get ahead of this yet
  * Add in a grouping category - multi state/multi-national dataset
  * Ttest / anova

**Type of Data:**
Age distributions
Specifically generate a dataset with old/young folks over-represented to highlight a bimodal distribution

Start with single state/country
Add a second state/country to demo ttest
Add more to demo anova

Alternatively, income by education level - may be more interesting/relevant to college students (or depressing)

## Intro to R/RStudio
## Reading Data / Distributions
### What *is* a **normal distribution**
#### How normal is it?
show increasingly unclear examples of normal vs not

introduce tests of normality

#### Measuring normality - single sample

reinforce [concept of statistical] **normality**

is a value from a sample? - one way ttest
something about tails

#### comparing normality - two saples

standard / two-way t test

#### comparing more than two - ANOVA

#Glossary
Data
Quantitative
Qualitative
Discrete 
Continuous
Nominal
Ordinal

