
# Week 2: Intro to R, data types, data structures 

**NEEDS WORK**

**How to make more engaging**






## Load the Data 

**From last week**

We'll load the same data as last week and inspect some elements.

*OR* We could collect NEW data for this lesson. *shrug*

```{r}

dir_path <- file.path("inst","unit1_data")
survey_path <- file.path(dir_path, "data_template.xlsx")

data <- readxl::read_excel(survey_path)
```





## Inspect the Data 

What is `data`?? Below we call the `class()` function on `data` and see that it has 3 classes: 
  `tbl_df` , `tbl` , `data.frame`
  
The first two classes, `tbl_df, tbl` indicate it is a special kind of table, in the `tibble` format. In general, you can interact with these like a `matrix` or `data.frame` but they have additional features. 


```{r}
class(data)
```



We can call `colnames()` on data, like a regular `data.frame` or `matrix`. Or we can take advantage of the `tibble` structure and use the `glimpse()` function which provides a succinct summary of your data.

```{r}

colnames(data)

tibble::glimpse(data)

```








