# Appendix 2: Intro to R/Rstudio {.unnumbered}




A crash course to teach you everthing you need to know...




To teach yourself more with the rest of this book!



## Basics {.unnumbered}

### R vs RStudio {.unnumbered}



Let's start with some analogies:

  * If `R` is the engine, then `RStudio` is the car.
  
  * If `R` is the text, `RStudio` is the text-editor.
  
  
    
    
**What do you think some differences are between R and RStudio?**

<details>
  <summary> *Click to show answer* </summary> <br> 
  <p> 
  
  >
  > `R` is technically the **programming language**, 
  > `RStudio` is the **Integrated Development Environment (IDE)**
  >
  
  
  </p>
</details>

`RStudio` is a program/app just like Google Chrome or Microsoft Word. Each
of these programs provide a **Graphical User Interface (GUI)**, a pretty way for 
a user to use a mouse and keyboard to do *something.*

An **IDE** is a special kind of program/app that provides MANY tools for writing and running code.

Throughout this text, we will use **RStudio**, to write (aka, **code**) **R-scripts** that will transform, analyze, and/or visualize data.

**R** is a language that can DO a lot.
**RStudio** is a program that lets **YOU** do a lot **USING** the **R language** (and other languages).



### RStudio vs POSIT {.unnumbered}


One last bit of definitions: **POSIT** is a new name/re brand of the company that produces and maintains the RStudio software. In the past, the RStudio (company) produced the RStudio (software).


Now, POSIT (company) produces RStudio (software), along with Quarto (software), and other resource for data scientists.





## How to install {.unnumbered}



R is available directly from the Comprehensive R Archive Network, CRAN, located at https://cran.r-project.org/. You can find the latest official and development versions of **base-R** and all contributed packages hosted by CRAN.

You can also go straight to the POSIT website, https://posit.co/download/rstudio-desktop/, where you can find download links for both R and Rstudio.




### Base R and beyond {.unnumbered}



**Base R** Generally refers to the core set of **R packages** that come bundled together when you download R from CRAN or POSIT. **Packages** are collections of R code that can expand the functionality that is not included with the base R packages. For example, this book was made using the `bookdown` Rpackage which itself extends the `Rmarkdown` package. Together, these packages make it easy to produce high quality textbook style materials using only R!.

This section will focus on using base R commands to orient users, however this book does draw on several additional packages, namely the IPUMS packages: 

  * ipumsr
  * ipumsED
  * ipumsEDbook (that's the text you're reading!)
  
  
  
R packages need to **installed** one time which downloads a copy to your local R package library. This is most often done using the `install.packages()` command, which whill download R packages straight from CRAN. 

For example, you could run the following to download the packages mentioned above. Note how the `c()` function is needed in order to provide a list of R packages the user wishes to download.

```{r, eval = FALSE}

## One package
install.packages("ipumsr")


## Several packages

install.packages(c("ipumsr", "ipumsED", "ipumsEDbook"))

```





Then, any time you want to use functionality from that package, you can use the `library()` function at the start of your **R session**. In general, your **R session** lasts for the entire time you are actively working. You can **restart** your session in order to clear your workspace, memory, etc. Or, if your computer goes to sleep, your session will also (most likely) automatically restart.


Note that `library()` does not expect a characer string (in quotes `""`), and that it only takes a single package name. 


```{r, eval = FALSE}

library(ipumsr)
library(ipumsED)

```




Note also that while downloading `ipumsEDbook` provides all the source code for this document, it is unlikely that you would ever need to load the package with `library()`. Instead, you can use the following notation to call a specific function without loading the entire package. In this case, we call a function which will regenerate the html version of this document. 




```{r, eval = FALSE}

ipumsEDbook::regenerate_book()

```

 
## Navigating RStudio {.unnumbered}


**Lots of Panes, not a pain!**

> You get used to it, promise.



### Navigating... Console

The console is where you interact with `R` directly. 

The console is where you actually input the code into R

#### Writing R code - give it a try

At its simplest, R is a calculator. You can enter basic math equations into the console like. R will read the input, interpret it, and print some kine of output.

The exact output you receive depends on the input!



```{r}

2+2

```



As a statistical programing language, R is capable of basic arithmetic and following order of operations. 




```{r}

(2+2)^2/2

```


#### R Objects

R is also an oject oriented programing language. This means you can create R-objects. These objects can store complicated input or output. R objects can really be anything, and R objects can be used to create new types, or **clases** of R objects.

Let's start simple. Try to describe, in plain language, what's going on in the code below. In other words, how do you as a human, interpret the code.

What do you think will happen when we **run the code**

```{r}

my_answer <- 1+2+3+4+5+6+7+8+9

```





<details>
 <summary> *Click to show answer* </summary> <br> 
  <p> 
  
  > 
  > In the code chunk above, we create a new R object called my_answer. This
  > object is the result or sum of the numbers 1 through 9.
  >
    
  </p>
</details>




So what *is* `my_answer`???

In the above code chunk, we add up the numbers 1 through 9 and save the result to a new R object, which we call `my_answer`. R will print results by default, as we saw in the first math expressions. But in the above codechunk, we "catch" the result, so to speak, in out R object. We can access the result by calling the R object by name.



```{r}

my_answer

```

#### R functions {.unnumbered}

In addition to objects, a second important concept to programing in R is **functions**. Functions are actually just a special **class** of R object. What makes functions special is that they **do** something. In gneral functions take one or more **arguments** as **input**, they do some kind of action like calculations, visualizations, or data transformations, and the produce some kind of **output**

R functions are easily recognizable because they end in parentheses `()`. Arguments to the function go within the parentheses.



Instead of typing the `+` sign multiple times, as we did above, we could use the `sum()` function. In the code on the right, we pass the numbers 1 though 9 as arguments to the `sum()` functiom. The arguments go within the `()` separated by commas.


.pull-left[
```{r}

1+2+3+4+5+6+7+8+9

```


]



.pull-right[

```{r}

sum(1,2,3,4,5,6,7,8,9)

```

]



When we create objects in R, we use `<-`, which is known as the **assignment operator**. The assignment operator is a special function. instead of using `()`, Everything to the right of the assignment operator is taken as the input. Whatever is to the left of the assignment operator is also input! This becomes the name of the R object. 


The output of the assignment operator is the R object itself. 

```{r}

my_sum <- sum(1,2,3,4,5,6,7,8,9)

my_sum

```



>
> Notice how we create the object, `my_sum` and then call the object, by name,
> on the next line, in order to print the results.
>


**Be careful!** The assignment operator **does** have some limitations on what it can/can't take as input. Try to run the following code in your own R session:

```{r, eval = FALSE}


my_sum <- 1,2,3,4,5

```




<details>
 <summary> *If you can't run the code, click to show result* </summary> <br> 
  <p> 
  
  > ```
  > Error: unexpected ',' in "my_sum <- 1,"
  > ```
  >
  
  </p>
</details>


So what went wrong??


Technically, the assignment operator is expecting just one thing to the right of it. In the working code `my_sum <- sum(1,2,3,4,5,6,7,8,9)` , R interprets, or evaluates, the `sum()` function, calculating a result that is a single value, `r sum(1:9)`, which we give the name `my_sum`.


If you want to store a list of values, you need to **concatenate** them first. This is probably a common concept to anyone with some kind of programing experience before. To concatenate something is simply to put it together. In `R`, you will get **very** used to using the concatenate function - in fact, it's so common the creators of R wisely gave it a very simple, easy to remember name: `c()`. 


Using `c()` is easy. It takes any number of inputs, and joins them together into a single **vector**. A **vector** is an R object that lists several values. Notice the difference?


.pull-left[
```{r}

my_value <- sum(1,2,3,4,5,6,7,8,9)

my_value
```

]


.pull-right[
```{r}
my_vector <- c(1,2,3,4,5,6,7,8,9)

my_vector
```

]


#### Interacting with Objects and Functions 

We can use some basic R functions to interact with our R objects. The first function tells us the `class` of our R object. While `length()` tells us how long the object is. Notice how `my_value` and `my_vector` have the same class, but different lengths. 



.pull-left[
```{r}

my_value <- sum(1,2,3,4,5,6,7,8,9)

class(my_value)
length(my_value)

```

]


.pull-right[
```{r}
my_vector <- c(1,2,3,4,5,6,7,8,9)

class(my_vector)
length(my_vector)
```

]


There is technically no end to the number of R object classes that can exist in R. That's because classes can be built upon and modified to suit specialized purporses. Creating your own object classes is an advanced topic, for now, you will be interacting with pretty basic classes of objects.


Below are several different vectors that represent some of the most basic object classes. Notice how these classes align closely with data classes.

```{r}

my_logical <- c(TRUE, TRUE, FALSE, TRUE, FALSE,TRUE,FALSE,FALSE, TRUE)

my_numeric <- c(1,2,3,4,5,6,7,8,9)

my_character <- c("a", "b", "c", "d", "e", "f", "g","h", "i")

```


A vector stores one or more values of the same type of data.


If we want to combine multiple vectors together, we can do that with the `data.frame()` function. This function takes multiple vectors as inputs and combines them together into a single object, as long as they're the same length!


.pull-left[

**Why does this data.frame fail??**

```{r}

my_df <- data.frame(
  "logical" = my_logical,
  "numeric" = my_numeric,
  "character" = my_character,
  "numbers" = c(2,4,6)
)

```

]


.pull-right[


```{r}

my_df <- data.frame(
  "logical" = my_logical,
  "numeric" = my_numeric,
  "character" = my_character
)

```


]


Notice the arguments we pass to `data.frame()`. Do you see how each piece fits together?

Can describe how each argument goes from input to output?


<details>
 <summary> *Click to show answer* </summary> <br> 
  <p> 
  
  > 
  > In this example, we pass `data.frame()` three arguments, each separated by a comma `,`. 
  > In each argument, we specify an R object to the right of an equal sign `=` and 
  > a character string within quotes `""` to the left of the equal sign. 
  >
  > The characters to the left of the equal sign become the **names** of each column. The R objects, specified verbatim, become the contents of each column. 
  >
    
  </p>
</details>



```{r}

knitr::kable(my_df)
```



You're bound to create **MANY** R objects throughout your work, and that can be a lot to keep track of. Fortunately, you don't need to track this all yourself - that's what the environment tab is for!


### Navigating ... Environments 

Finally, we come to the second tab of Rstudio. In addition to Environment, you may see tabs for `History, Connections`. You may also see tabs for `Build` or `Git` depending on if you have these tools installed. 

While you *may* interact with these other tabs depending on your project needs, you will **always** want to reference R objects stored in your **local environment**. In the environment panel, you will find each R object listed by name and organized broadly into categories of `Values` (vectors), `Data` (data frames, matrices, and other more complex structures), User-created `Functions` will also recieve their own section. 


Your local environent is a temporary space. When you close your R session, you will lose any work you do not save or **write out** to a separate file. 



Of course, that doesn't mean you can't re-run old code, and it doesn't mean that you can't save your progress in between sessions. In fact, there are **Several** options for how RStudio handles your environment. By default, R will attempt to preserve a copy of all data in your local environement. It will also try to restore whatever you were working on most recently when you open a new session.


These may seem like handy convenience features, but in practice we find them prone to cause issues (outlined below).



#### Our Reccomendations

Here we outline what we consider to be best practice for maintaining an R environment/ R project.

##### CHANGE GLOBAL OPTIONS


For the most part, we strive to use the default installations/parameters when using R/Rstudio. This facilitates reproducibility for a researcher working across different computers, and reproducibility if a colleague wishes to replicate the code themselves. 

However, one setting we **DO** reccomend changing the following under `Tools/Global Options/General` Uncheck all boxes that begin `Restore`..

 * Restore last R project
 * Restore previous source files
 
 * Restore .RData
 
Also, we reccomend changing the option for `Save a copy of .RData` to `NEVER`.

The reason for this is that if an R session crashes, trying to load the latest autogenerated backup is likely to cause the error and make the session fail again. It can also be timely if you last worked on a really large project that takes a while to open and restore all files. 

In practice, it's better to intentionally save and close your project. Then open up whichever project you want to work on each time you start a new R session. 




### Writing R in Scripts (Breaking the 4th Pane)



Save your work, find typos, re-run easy!


## R-specific Quirks {.unnumbered}

### R objects, Functions {.unnumbered}

R is an object-oriented programming language. Everything in R is either an object or a function. R objects generally store or define some kind of data. R functions, generally, take R objects as input, do some kind of operation, and produce some kind of output. 

In the context of **language**, objects are the nouns and functions are the verbs.

When you write R code, you usually create R objects that represent data of interest, then apply various R functions to transform, analyze, or visualize the data. In the below code, we are creating an **object** called `my_data`, which represents the integers one through 10. We are using the `c()` **function** which takes any number of input **objects** as **input** and **concatenates** them together into a character string. 

The **input(s)** to a function go within the parentheses seperated by commas. You can usually tell the difference between R obects and R functions because R functions always have `()` written after their name.

In most cases, the **output** of an R function will be printed to the console, unless it is saved into a new R object. We do this using the **assignment opperator**, `<-`. This is a special function R function that takes whatever is on the right-hand side of the `<-` and creates or overwrites an R object with the name specified to the left-hand side of the `<-`. In the example below, we use the **assignment opperator** to "catch" the result of the `c()`, which is the character string of intengers from 1 to 10.


Calling the name of an R object will *usually* print its contents in the console.



```{r}

my_data <- c(1,2,3,4,5,6,7,8,9,10)



my_data

```



We say usually because objects have different properties depending on their **class**. 



> NOTE: We sometimes use the term **class** to refer to the class of data of something. Is a given number sequence comprised of integers, factors, or real numbers? In R, and in this text, when we use **class** we are explicitly referring to the R-object-class of an object. Because R is a statistical programming language, the data-class of an object and the object-class of an object are sometimes synonymous (EG factors, logical, character), other times, the names may be slightly different (EG the R-class numeric comprises both integer numbers and complex numbers).



### R classes

R classes determine the characteristics of the R object. They may set parameters about what kind of data can/can't be included in the R object. They may also define or modify preset ways these R objects interact with other R objects. You can use the `class()` function on any R object to find out it's class(es). 

That's right, an R object can have more than one class.

Some basic R classes are **vectors**, **data,frames**, and **functions**. That's right! a function is actually just a certain **class** of R object. You can see below that in our example, `my_data` hass the class **vector**.

```{r}

class(my_data)

```

### R Projects


R projects are an essential way to organize your data analysis, research, or even scratch work in R. Projects let you create self-contained environments that can feature their own R libraries, containg packages that are different from your day-to-day operations. 

### R code vs R scripts vs R markdown

R **code** can go anywhere. It gets input into the console to actually run the commands. But if you want to keep a record of your code and document it, you probably want to write an **R script**. A script is simply a text file, that contains code. Presumably, the code takes some inputs, does some operations, and produces helpful output. 

Sometimes we use scripts as a start-to-finish record of what we did in our work. This is often the case for research projects. A script might read data from a source, analyze it, visualize it, and save out results.

If we ever need to re-do this work, it's as simple as "re-runing" the script (literally a click of a button).

Scripts can also be helpful to automate workflows. If the same data transofrmations need to occur across many samples, a script can be written to **iterate** across a set of inputs, perform the same calculations/transformations, and output results in some way.

R markdown is just another script, but has a few more bells and whistles that make writing and documenting code a much more enjoyable experience. In fact, this whole book was written using Rmarkdown!!

## How to find help {.unnumbered}

### In R

All R functions *should* come with their hown help file. There are a few different ways to access this page.

In RStudio, you can click on the Help tab of the File/Plots/Packages...(**HELP**, Viewer, Presentation) panel - the Lower-Right panel by default. 

Here the serach box will search the help files for any R package you have installed. 

You could also access the help files by pressing the `F1` key while your cursor is at the end of an r function.

*todo: Screenshot of cursor placement after function name, before open parentheses*

### the web!

Google (or duck-duck-go or, i guess, bing) it!

"How to calculate area under the curve in R"

You may be directed directly to an R package tailor made for that purpose. In this case, the [AUC package](https://search.r-project.org/CRAN/refmans/DescTools/html/AUC.html) was the first result from the official CRAN page.

You can search CRAN if you know you want a package to do a particular thing. 

Alternatively, doing an internet search for a specific error message will often yield results on stackexchange or stackoverflow. These are two public sites where users submit responses and upvote helpful responses while downvoting unhelpful ones.

NOTE: posting directly to these sites CAN be helpful, but oftentimes someone has already asked the question and doing an internet search will get you the answer more quickly.




In general, people don't put out bad or malicious information about R, so you can trust any reasonable responses from the above websites. In addition, there are a few notable people or teams who consistently put out good (read amazing) R-related tips/help.


RStudio (Posit)
Yihui

Tidyverse
  Hadley Wickam
  Jenny Bryan
  




## Example Workflow {.unnumbered}


### Create a new R project
### Load a dataset
### Visualize/Analyze data
### Save out results

## Excercises {.unnumbered}


### Interpreting R

Describe what the following R code does, in plain words.

```{r, eval = FALSE, echo = TRUE}

my_data <- read_excel("//filepath/directory/filename.xlsx")

```


*Hint:* There are 4 elements in the above code: 
  `my_data`, 
  `<-`, 
  `read_excel()`, 
  `//filepath/directory/filename.xlsx`



### Try setting up a project

* Read data in from outside the R project
* Save data within your R project
* Save data external to your R project
* Visualize a dataset or trend.

 
