
# Appendix 3b: Additional lessons / content to transfer over: {.unnumbered}

## Intro to R, data types, data structures {.unnumbered}


In the previous lesson, we began thinking about **data**, how to talk about it, 
and how to **visualize** it. We also talked about one type of average, the **mean.**

## Engage {.unnumbered}

*What do you think the following code does?*


```{r, eval = FALSE, echo = TRUE}

my_data <- read_excel("//filepath/directory/filename.xlsx")

```



*Hint:* There are 4 "things" in the above code: 
  `my_data`, 
  `<-`, 
  `read_excel()`, 
  `//filepath/directory/filename.xlsx`



### R vs RStudio

* If `R` is the engine, then `RStudio` is the car.

* If `R` is the text, Rstudio is the text-editor.

* `R` is the **programming language**, `Rstudio` is the **Integrated Development Environment (IDE)**.
  * *What do you think some differences are between R and RStudio?*


`RStudio` is a program/app just like Google Chrome or Microsoft Word. Each
of these programs provide a **Graphical User Interface (GUI)**, a pretty way for 
a user to use a mouse and keyboard to do *something.*

An **IDE** is a special kind of program/app that provides MANY tools for writing and running code.



### Orientation to RStudio

_SUGGESTION: Instructor live demos interacting with RStudio while students follow along on computers_



When you first open RStudio, you'll see 3 **panes**, all of which will look fairly empty at the moment.

If you're using the default layout, you should see: 
  * On the left, the `Console` **pane**
  * On the Top-right, the `Environment` **pane**
  * On the Bottom-right, the `Files` **pane**


A keen eye will also notice that each of these **panes** contains multiple **tabs**. We will go over the uses of many of these **tabs**, but for now let's start with `Console.`



The `Console` is where you input `R code`, run it, and see the results.


At it's simplest `RStudio` is a calculator. Try typing `4 + 4` into the `Console`,
then press the `[Enter]` key to run the code. Immediately, `R` prints the result as we see here:

```{r}
4 + 4
```



The `Console` serves as a running log of all your operations for your current session, but it can be helpful to temporarily save your results as `R objects`. We do this using the  **assignment operator** we saw earlier: `<-`


```{r}
answer <- 4+4
```


By default, `Console` only prints results if they have no where else to go. If they're stored as an `Robject`, no result is printed, but you should now see `answer` listed in the `Environment` **pane**.





## Explore - Interacting with R objects  {.unnumbered}



### Load the Data 



```{r}

dir_path <- file.path("inst","unit1_data")
survey_path <- file.path(dir_path, "data_template.xlsx")

data <- readxl::read_excel(survey_path)
```






What is `data`?? Below we call the `class()` function on `data` and see that it has 3 classes: 
  `tbl_df` , `tbl` , `data.frame`
  
The first two classes, `tbl_df, tbl` indicate it is a special kind of table, in the `tibble` format. In general, you can interact with these like a `matrix` or `data.frame` but they have additional features. 


```{r}
class(data)
```



We can call `colnames()` on data, like a regular `data.frame` or `matrix`. Or we can take advantage of the `tibble` structure and use the `glimpse()` function which provides a succinct summary of your data.

```{r}

colnames(data)

tibble::glimpse(data)

```





### Inspect the Data 

What is `data`?? Below we call the `class()` function on `data` and see that it has 3 classes: 
  `tbl_df` , `tbl` , `data.frame`
  
The first two classes, `tbl_df, tbl` indicate it is a special kind of table, in the `tibble` format. In general, you can interact with these like a `matrix` or `data.frame` but they have additional features. 


```{r}
class(data)
```



We can call `colnames()` on data, like a regular `data.frame` or `matrix`. Or we can take advantage of the `tibble` structure and use the `glimpse()` function which provides a succinct summary of your data.

```{r}

colnames(data)

tibble::glimpse(data)

```





## Summarize Data {-}

### Continuous Data {-}

For continuous data, we often want to summarize our data by describing the **mean, median, and/or range**. **Mean** and **median** describe the *central tendency* of the data, while **range** describes the full extant of the data, as seen below. 

NOTE: if `NA` are present in the data, be sure to use the `na.rm=TRUE` flag for these operations.


```{r}

mean(data$Height_inches, na.rm = T)
median(data$Height_inches, na.rm = T)
range(data$Height_inches, na.rm = T)

```




### All in summary() {-}


**Mean**, **median**, and **range** will all be reported by calling `summary()` on a `numeric vector`, such as `Height_inches`. In addition, the lower and upper quartiles will be reported, along with the number of `NA` responses.


NOTE: `summary()` does NOT require special handling for `NA` values, in fact - it expects them!

```{r}
summary(data$Height_inches)
```


## Mode {.unnumbered}

You're probably familiar with **mean** and **median** being talked about with a 
third term, **mode**. The **mode** is the most commonly occuring value in a 
dataset. It's often important to know the **modal response** of survey data. 
While a commonly reported metric(??), there is no `mode()` function included in 
`base r`...  




so we'll just have to create our own!   



### Mode Code 

One common measure of data reported is the mode, or most frequently occuring value. For whatever reason, this is not a default function in R, but we can easily write our own function like so:

```{r}

my_mode <- function(x){
  tt <- table(x) ## find frequencies
  tt <- tt[order(tt, decreasing = TRUE)] ## resort based on freq
  
  ## check number of modes
  max <- max(tt)
  n_max <- sum(tt==max)
  
  
  if(n_max > 1 ){
    warning("More than one mode detected")
    return(tt[tt==max])
  } else {
    ## return only the first value
    return(tt[1]) ## return whatever the highrst frequency is
  }

  
}

```

### Mode Results  


Now that we've created out own `function`, it's easy to find the **mode**


```{r}

my_mode(data$Height_inches)

```



## Visualizing Data {.unnumbered}

The above summaries describe data with numbers, but we can also describe data visually.



### Continuous Data - Boxplots 


Univariate continuous data, like height, can be visualized using a box and whisker plot, which shows many of the components of summary:

  * the **median** is the black bar in the middle
  * the **quartiles** (25th and 75th percentiles) are represented by the extents of the boxes
  * The **range** is shown by the whiskers, with outliers shown indvidually, if needed.
  
  
  
```{r}
boxplot(data$Height_inches)
```


### Continuous Data - Histograms 

Continuous data, can also be broken into **bins** and plotted as a **histogram**. The `hist()` function will attempt to find the optimum number of bins for you, but you can specify a different number with the `breaks` argument.


```{r}
hist(data$Height_inches, main = "Histogram of Height", xlab = "Height (in)")
```




### Categorical Data 

Categorical data is already in discrete units. In general with categorical data, we want to count the **frequency** of unique values. There are many ways to do this, but one of the easiest is the `table()` function. Saving the results of the table to an object, `birth_freq`, allows you to save and print the results at any time.

```{r}

birth_freq <- table(data$Birth_Month)

birth_freq
```


We can also visualize our tabulated results using a **barplot** as below.


```{r}

barplot(birth_freq)

```



## Glossary {.unnumbered}


